---
title: "Assignment 2b: Evaluating Classification Model Performance"
author: "Kevin Villa"
date: "`r Sys.Date()`"
output:
  html_document: default
---


## Introduction

We are evaluating a binary classification model that predicts penguin
sex. The dataset (`penguin_predictions.csv`) has the model's predicted probability
of "female," its default predicted class, and the actual sex label. The
goal is to understand how changing the probability threshold affects model
performance.

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Body

```{r load-data}
penguins <- read_csv("https://raw.githubusercontent.com/acatlin/data/master/penguin_predictions.csv")
glimpse(penguins)
```

```{r null-error-rate}

class_counts <- count(penguins, sex)
class_counts

majority_count <- max(class_counts$n)
total <- nrow(penguins)
null_error_rate <- 1 - (majority_count / total)
null_error_rate
```

```{r class-distribution}
ggplot(penguins, aes(x = sex)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Actual Sex Class", x = "Sex", y = "Count") +
  theme_minimal()
```

```{r confusion-matrices}

threshold_results <- function(threshold) {
  pred <- ifelse(penguins$.pred_female > threshold, "female", "male")

  TP <- sum(pred == "female" & penguins$sex == "female")
  FP <- sum(pred == "female" & penguins$sex == "male")
  TN <- sum(pred == "male"   & penguins$sex == "male")
  FN <- sum(pred == "male"   & penguins$sex == "female")

  tibble(threshold = threshold, TP = TP, FP = FP, TN = TN, FN = FN)
}

thresholds <- c(0.2, 0.5, 0.8)
confusion_table <- bind_rows(lapply(thresholds, threshold_results))
confusion_table
```

```{r pretty-confusion-matrices}

for (t in thresholds) {
  pred <- ifelse(penguins$.pred_female > t, "female", "male")

  cm <- matrix(
    c(
      sum(pred == "female" & penguins$sex == "female"),
      sum(pred == "male"   & penguins$sex == "female"),
      sum(pred == "female" & penguins$sex == "male"),
      sum(pred == "male"   & penguins$sex == "male")
    ),
    nrow = 2,
    dimnames = list(
      "Predicted" = c("Female", "Male"),
      "Actual"    = c("Female", "Male")
    )
  )

  print(paste("Threshold:", t))
  print(kable(cm))
}
```

```{r metrics}

metrics_table <- confusion_table %>%
  mutate(
    accuracy  = (TP + TN) / (TP + FP + TN + FN),
    precision = TP / (TP + FP),
    recall    = TP / (TP + FN),
    f1        = 2 * TP / (2 * TP + FP + FN)
  )

metrics_table
```

## Conclusions

The null error rate for this dataset is 41.9% always guessing "male"
(the majority class) would be wrong nearly 42% of the time. Every threshold
tested dramatically outperforms this baseline, with accuracy ranging from
91.4% (threshold 0.2) to 94.6% (threshold 0.8), confirming the model carries
genuine predictive signal rather than just reflecting class imbalance.

As the threshold increases from 0.2 to 0.8, precision rises steadily (0.860
to 0.947) while recall stays roughly flat (0.949 to 0.923), meaning higher
thresholds mostly filtered out false positives here without costing many
true positives, which is why accuracy and F1 both peaked at the highest
threshold (0.8) for this particular dataset.

**A 0.2 threshold** is better when missing a true positive is costly and
false alarms are cheap to double check, for example, if researchers are
using this model to flag candidate female penguins for a follow-up field
study, they'd rather over flag and manually verify a few extra birds than
miss real females entirely.

**A 0.8 threshold** is preferable when a false positive carries real
downstream cost, for example, if the sex prediction feeds directly into a
genetics or population dataset without human review, a wrongly-labeled
penguin could quietly corrupt that dataset, so it's worth being more
conservative before committing to a "female" label.