---
title: "Assignment 2: SQL and R - Album Ratings"
author: "Kevin"
date: "`r Sys.Date()`"
output:
  html_document: default
---

## Introduction

This analysis collects listener ratings (1-5 scale) for six tracks from The
Strokes' album *Reality Awaits*, stores them in a SQLite database using a
normalized three-table schema, and loads the results into R for analysis.

```{r setup, include=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)
```

## Body

```{r create-db}
con <- dbConnect(RSQLite::SQLite(), "album_ratings.db")

dbExecute(con, "DROP TABLE IF EXISTS ratings;")
dbExecute(con, "DROP TABLE IF EXISTS songs;")
dbExecute(con, "DROP TABLE IF EXISTS users;")

# Three-table schema: users and songs are independent entities,
# ratings is the junction table linking them (many-to-many relationship)
dbExecute(con, "
CREATE TABLE users (
  user_id INTEGER PRIMARY KEY,
  name TEXT NOT NULL
);")

dbExecute(con, "
CREATE TABLE songs (
  song_id INTEGER PRIMARY KEY,
  title TEXT NOT NULL
);")

dbExecute(con, "
CREATE TABLE ratings (
  user_id INTEGER,
  song_id INTEGER,
  rating INTEGER CHECK (rating BETWEEN 1 AND 5),
  PRIMARY KEY (user_id, song_id),
  FOREIGN KEY (user_id) REFERENCES users(user_id),
  FOREIGN KEY (song_id) REFERENCES songs(song_id)
);")
```

```{r populate-db}
dbExecute(con, "INSERT INTO users (user_id, name) VALUES
  (1, 'Person 1'), (2, 'Person 2'), (3, 'Person 3'), (4, 'Person 4'), (5, 'Person 5');")

dbExecute(con, "INSERT INTO songs (song_id, title) VALUES
  (1, 'Dine and Dash'), (2, 'Lonely in the Future'), (3, 'Going to Babylon'),
  (4, 'Going Shopping'), (5, \"Liar's Remorse\"), (6, 'Psycho Shit');")

# Person 3 has no rating for song 5, Person 5 has none for song 4 --
# these gaps simulate real survey conditions where not everyone rates everything
dbExecute(con, "INSERT INTO ratings (user_id, song_id, rating) VALUES
  (1,1,4), (1,2,5), (1,3,5), (1,4,3), (1,5,3), (1,6,2),
  (2,1,3), (2,2,4), (2,3,5), (2,4,3), (2,5,2), (2,6,2),
  (3,1,4), (3,2,5), (3,3,3), (3,4,3),          (3,6,2),
  (4,1,3), (4,2,5), (4,3,3), (4,4,3), (4,5,2), (4,6,2),
  (5,1,3), (5,2,4), (5,3,3),          (5,5,3), (5,6,3);")
```

```{r query-data}
# CROSS JOIN every user against every song, then LEFT JOIN actual ratings --
# any user/song pair with no matching rating row surfaces as NULL/NA here,
# which is how we detect and preserve missing data rather than hiding it
ratings_full <- dbGetQuery(con, "
  SELECT u.name AS listener, s.title AS song, r.rating
  FROM users u
  CROSS JOIN songs s
  LEFT JOIN ratings r
    ON r.user_id = u.user_id AND r.song_id = s.song_id
  ORDER BY u.name, s.title;
")

glimpse(ratings_full)
```

```{r summarize}
# A missing rating means the listener hasn't reviewed that track -- not that
# they rated it poorly. We keep it as NA and exclude it (na.rm = TRUE) from
# every summary stat, rather than guessing a value or treating it as zero.

song_summary <- ratings_full %>%
  group_by(song) %>%
  summarize(
    avg_rating = round(mean(rating, na.rm = TRUE), 2),
    num_ratings = sum(!is.na(rating)),
    num_missing = sum(is.na(rating))
  )

song_summary
```

```{r cleanup}
dbDisconnect(con)
```

## Conclusions

Across all six tracks, "Lonely in the Future" was the clear favorite at an
average rating of 4.6, while "Psycho Shit" and "Liar's Remorse" scored lowest
at 2.2 and 2.5 respectively. Two ratings were missing out of 30 possible
(Person 3 never rated "Liar's Remorse," Person 5 never rated "Going
Shopping"), and both were excluded from the averages rather than imputed,
since a missing rating reflects "hasn't listened" rather than a neutral or
negative opinion.

Future work could expand the listener pool beyond five people, explore
whether ratings correlate with track position on the album, or normalize
each listener's ratings to account for people who rate more harshly or
generously overall.