---
title: "Project 2 — Data Tidying: Life Expectancy by Country"
author: "Kevin Villa"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
---

## 4. Data Source

This dataset is the World Bank's "Life expectancy at birth, total (years)" indicator (`SP.DYN.LE00.IN`), pulled from the World Bank API for 16 countries across 2000–2023. This is the same dataset a classmate picked for the Week 5 Discussion 5A post: each country is a row, and each year is stored as its own column, so year — which is really a variable — is spread across 24 separate column headers instead of living in one column.

Source: World Bank, "Life expectancy at birth, total (years)," `https://data.worldbank.org/indicator/SP.DYN.LE00.IN`.

The raw file, preserving the original wide (country-by-year) structure with no tidying applied, is committed here:

`https://raw.githubusercontent.com/nowhyporque/data607DataAcquisitionAndManagement/refs/heads/main/Project2/life_expectancy_raw.csv`

```{r setup}
#| label: setup
#| include: false
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(knitr)
```

## 5. Data Structure Before Tidying

```{r load-raw}
#| label: load-raw
life_raw <- read_csv("https://raw.githubusercontent.com/nowhyporque/data607DataAcquisitionAndManagement/refs/heads/main/Project2/life_expectancy_raw.csv")

glimpse(life_raw)
kable(life_raw[, 1:6])
```

There are 16 rows (one per country) and 25 columns (`Country` plus one column per year, 2000 through 2023). Year is a variable baked into the column names rather than stored as values, which is exactly the untidy pattern described in the discussion post. This particular slice of the data has no missing cells, but the transformation step below still documents how a gap would be handled if one existed.

## 6. Transformation Steps

```{r tidy-pipeline}
#| label: tidy-pipeline
# Reshape from wide to long: year becomes a variable in its own column
# instead of 24 separate year columns.
life_long <- life_raw %>%
  pivot_longer(
    cols = -Country,
    names_to = "year",
    values_to = "life_expectancy"
  ) %>%
  # Normalize variable structure / rename to a consistent naming
  # convention: year as an integer rather than text, lowercase column names
  mutate(
    year = as.integer(year),
    country = Country
  ) %>%
  select(country, year, life_expectancy) %>%
  arrange(country, year)

# Missing values: none of these 16 countries have a gap in 2000-2023, but
# if a country/year combination were missing, it would surface here as NA
# after pivoting rather than silently disappearing, so I'm flagging the
# handling decision explicitly: any NA is left as NA (not imputed to 0 or
# interpolated), since fabricating a life expectancy value would be
# misleading, and simply omitted from any average calculated below.
sum(is.na(life_long$life_expectancy))

kable(head(life_long, 8))
```

## 7. Analytical Methods

The discussion post's business questions were about comparing life expectancy across countries over time, finding which countries improved the most, and looking at the COVID-19 dip and recovery. I calculate the change in life expectancy from 2000 to 2023 for each country, and separately look at the dip from 2019 to 2021 (the pandemic years) and the recovery by 2023.

```{r analysis}
#| label: analysis
change_2000_2023 <- life_long %>%
  filter(year %in% c(2000, 2023)) %>%
  pivot_wider(names_from = year, values_from = life_expectancy, names_prefix = "yr_") %>%
  mutate(change = round(yr_2023 - yr_2000, 2)) %>%
  arrange(desc(change))

kable(change_2000_2023, caption = "Change in life expectancy, 2000 to 2023")

covid_impact <- life_long %>%
  filter(year %in% c(2019, 2021, 2023)) %>%
  pivot_wider(names_from = year, values_from = life_expectancy, names_prefix = "yr_") %>%
  mutate(
    covid_dip = round(yr_2019 - yr_2021, 2),
    recovery_by_2023 = round(yr_2023 - yr_2019, 2)
  ) %>%
  arrange(desc(covid_dip)) %>%
  select(country, yr_2019, yr_2021, covid_dip, yr_2023, recovery_by_2023)

kable(covid_impact, caption = "COVID-19 dip (2019-2021) and recovery by 2023")
```

```{r chart}
#| label: chart
ggplot(life_long, aes(x = year, y = life_expectancy, color = country)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Life Expectancy by Country, 2000-2023",
       x = "Year", y = "Life Expectancy (years)", color = "Country") +
  theme_minimal()
```

India improved the most over the full period, gaining 9.25 years of life expectancy from 2000 to 2023, followed by Kenya (+7.57) and South Korea (+7.52). Every country in the dataset shows a visible dip around 2020-2021 on the chart, but the size of the dip varies a lot: Mexico lost 4.78 years between 2019 and 2021, the worst in this dataset, while higher-income countries like Canada and Japan barely dipped at all. By 2023 most countries had mostly recovered, though Mexico's rebound (+0.54 years from 2019 to 2023) is still far short of fully making up its pandemic-era loss.

## 8. Conclusions

Reshaping year out of the column headers and into its own variable made it possible to plot a proper time trend and calculate year-over-year changes, neither of which would be straightforward with year baked into 24 separate columns. The long-run trend is one of steady improvement almost everywhere, but the sharp, uneven COVID-era dip shows that "steady" isn't the same as "guaranteed" — a few years of a shared global shock hit some countries (Mexico, India, Indonesia) far harder than others (Canada, Japan, China), and by 2023 not every country had fully recovered to its pre-pandemic trajectory. A natural next step would be to pull in a wider set of countries and correlate the size of each country's COVID-era dip with factors like healthcare spending or vaccination rates.
