---
title: "Project 2 — Data Tidying: NFL Touchdown Scoring by Type"
author: "Kevin Villa"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
---

## 4. Data Source

This dataset is the official NFL.com team scoring table for the 2026 regular season, pulled from `https://www.nfl.com/stats/team-stats/offense/scoring/2026/reg/all`. It's the same kind of dataset Daanish picked for the Week 5 Discussion 5A post: a team-by-team breakdown of how each team's touchdowns were scored (rushing vs. receiving vs. two-point conversions), with the touchdown type spread across separate columns instead of stored as a variable.

The raw file, exactly as scraped from the site with no tidying applied, is committed here:

`https://raw.githubusercontent.com/nowhyporque/data607DataAcquisitionAndManagement/refs/heads/main/Project2/nfl_scoring_raw.csv`

```{r setup}
#| label: setup
#| include: false
library(readr)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(knitr)
```

## 5. Data Structure Before Tidying

```{r load-raw}
#| label: load-raw
nfl_raw <- read_csv("https://raw.githubusercontent.com/nowhyporque/data607DataAcquisitionAndManagement/refs/heads/main/Project2/nfl_scoring_raw.csv")

glimpse(nfl_raw)
kable(head(nfl_raw, 8))
```

This is wide and untidy for the same reason as the example in the assignment: touchdown *type* is a variable, but instead of living in its own column, it's spread across the column headers `Rsh_TD`, `Rec_TD`, and `Two_PT`. There are 32 rows (one per team) and 5 columns. `Tot_TD` is a pre-computed total rather than a raw observation, so it isn't itself a touchdown type — it gets handled separately in the transformation step below.

## 6. Transformation Steps

```{r tidy-pipeline}
#| label: tidy-pipeline
# Reshape from wide to long: touchdown type becomes a variable instead of
# three separate columns. Tot_TD is excluded here since it's a derived
# total, not a touchdown-type observation — it gets recomputed after
# reshaping instead, so it always agrees with the long-format data.
nfl_long <- nfl_raw %>%
  select(-Tot_TD) %>%
  pivot_longer(
    cols = c(Rsh_TD, Rec_TD, Two_PT),
    names_to = "td_type",
    values_to = "count"
  ) %>%
  # Normalize variable structure / rename to a consistent naming
  # convention: lowercase, descriptive labels instead of abbreviations
  mutate(
    td_type = case_when(
      td_type == "Rsh_TD" ~ "rushing",
      td_type == "Rec_TD" ~ "receiving",
      td_type == "Two_PT" ~ "two_point_conversion"
    ),
    team = str_to_lower(Team)
  ) %>%
  rename(team_name = Team) %>%
  select(team_name, team, td_type, count)

# Missing/inconsistent values: every cell in this source table is a real
# reported number (0 is a valid count, not a missing value — e.g. the
# Packers really did score zero rushing touchdowns), so there's nothing to
# impute here. The code below still documents the handling decision by
# explicitly treating any truly missing cell as 0 touchdowns, since a blank
# in a scoring table means "none scored," not "unknown."
nfl_long <- nfl_long %>%
  mutate(count = replace_na(count, 0))

kable(head(nfl_long, 9))
```

## 7. Analytical Methods

The business question from the discussion post was to compare each team's *scoring style* — the ratio of touchdowns by type, not just raw counts. I calculate each team's rushing and receiving touchdowns as a percentage of total touchdowns (excluding two-point conversions, which aren't full touchdowns), then rank teams by how rush-heavy or pass-heavy their scoring is.

```{r analysis}
#| label: analysis
scoring_style <- nfl_long %>%
  filter(td_type %in% c("rushing", "receiving")) %>%
  group_by(team_name) %>%
  mutate(total_td = sum(count)) %>%
  ungroup() %>%
  filter(total_td > 0) %>%
  mutate(pct_of_td = round(count / total_td * 100, 1))

rushing_share <- scoring_style %>%
  filter(td_type == "rushing") %>%
  select(team_name, total_td, pct_rushing = pct_of_td) %>%
  arrange(desc(pct_rushing))

kable(head(rushing_share, 5), caption = "Most run-heavy scoring teams")
kable(tail(rushing_share, 5), caption = "Most pass-heavy scoring teams")
```

```{r chart}
#| label: chart
ggplot(scoring_style, aes(x = reorder(team_name, pct_of_td), y = pct_of_td, fill = td_type)) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  labs(title = "Share of Touchdowns by Type, All 32 Teams",
       x = "Team", y = "Percent of Touchdowns", fill = "TD Type") +
  theme_minimal(base_size = 8)
```

The league as a whole leans pass-heavy: 62.2% of all touchdowns this season came through the air versus 35.0% on the ground (the remaining share is two-point conversions). But individual teams vary a lot. The Dolphins are the most run-heavy scoring team, with 75% of their touchdowns coming on the ground, while the Packers are the opposite extreme — 100% of their touchdowns this season have been through the air, with zero rushing touchdowns.

## 8. Conclusions

Reshaping this table from wide to long made it possible to calculate a real rate (percent of touchdowns by type) instead of just comparing raw counts, which would have been misleading since teams have scored different total numbers of touchdowns. The spread between the most run-heavy team (Dolphins, 75%) and the most pass-heavy team (Packers, 100% through the air) shows real differences in offensive identity across the league, not just differences in how many points teams have scored overall. A natural next step would be to track this breakdown week by week to see whether a team's scoring style shifts over the course of the season, for example if a run-heavy team becomes more pass-reliant once trailing in the standings.
