---
title: "Assignment 3b: Window Functions"
author: "Kevin Villa"
date: "`r Sys.Date()`"
output:
  html_document: default
---


## Introduction

This analysis generates a simulated daily price dataset for three
instruments -- Apple (AAPL), Microsoft (MSFT), and Bitcoin (BTC) -- from
January 1, 2022 through December 31, 2023, and uses SQL window functions
to calculate a year-to-date running average and a 6-day moving average
for each instrument.

```{r setup, include=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
```

## Body

```{r generate-data}

set.seed(607)

dates <- seq(as.Date("2022-01-01"), as.Date("2023-12-31"), by = "day")
n <- length(dates)

simulate_price <- function(start_price, drift, volatility, n) {
  changes <- rnorm(n, mean = drift, sd = volatility)
  pmax(1, round(start_price + cumsum(changes), 2))  
}

prices <- bind_rows(
  tibble(ticker = "AAPL", date = dates, price = simulate_price(175,    0.05, 2,   n)),
  tibble(ticker = "MSFT", date = dates, price = simulate_price(310,    0.08, 3,   n)),
  tibble(ticker = "BTC",  date = dates, price = simulate_price(46000,  5,    800, n))
)

glimpse(prices)
```

```{r load-into-sqlite}
con <- dbConnect(RSQLite::SQLite(), "prices.db")


dbWriteTable(con, "prices", prices, overwrite = TRUE)

dbGetQuery(con, "SELECT COUNT(*) AS row_count FROM prices;")
```

```{r window-functions}

window_results <- dbGetQuery(con, "
  SELECT
    ticker,
    date,
    price,
    ROUND(AVG(price) OVER (
      PARTITION BY ticker, strftime('%Y', date)
      ORDER BY date
      ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW
    ), 2) AS ytd_avg,
    ROUND(AVG(price) OVER (
      PARTITION BY ticker
      ORDER BY date
      ROWS BETWEEN 5 PRECEDING AND CURRENT ROW
    ), 2) AS moving_avg_6day
  FROM prices
  ORDER BY ticker, date;
")

glimpse(window_results)
head(window_results, 10)
tail(window_results, 10)
```

```{r plot}
window_results$date <- as.Date(window_results$date)

ggplot(window_results, aes(x = date)) +
  geom_line(aes(y = price, color = "Daily Price"), alpha = 0.3) +
  geom_line(aes(y = moving_avg_6day, color = "6-Day Moving Avg"), linewidth = 0.6) +
  geom_line(aes(y = ytd_avg, color = "YTD Avg"), linewidth = 0.6) +
  facet_wrap(~ ticker, scales = "free_y") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "Price, 6-Day Moving Average, and YTD Average by Instrument",
       x = "Date", y = "Price", color = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r cleanup}
dbDisconnect(con)
```

## Conclusions

*(To fill in once the code runs and the plot renders -- will summarize
how closely the 6-day moving average tracks daily price swings versus
the smoother, slower-moving YTD average, and note any differences in
volatility across AAPL, MSFT, and BTC.)*