---
title: "Assignment 3A: Global Baseline Estimates"
author: "Kevin Villa"
date: "`r Sys.Date()`"
output:
  html_document: default
---

## Introduction

This analysis uses a Global Baseline Estimate recommender system,
which is a non-personalized recommendation algorithm, using listener rating data for six songs from The Strokes' album *Reality Awaits*. The goal is to
predict what a listener would likely rate a song they haven't heard yet,
using only the overall average rating, that listener's personal rating
tendency, and that song's overall rating tendency.

```{r setup, include=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)
```

## Body

```{r create-db}
con <- dbConnect(RSQLite::SQLite(), "album_ratings.db")

dbExecute(con, "DROP TABLE IF EXISTS ratings;")
dbExecute(con, "DROP TABLE IF EXISTS songs;")
dbExecute(con, "DROP TABLE IF EXISTS users;")

dbExecute(con, "
CREATE TABLE users (
  user_id INTEGER PRIMARY KEY,
  name TEXT NOT NULL
);")

dbExecute(con, "
CREATE TABLE songs (
  song_id INTEGER PRIMARY KEY,
  title TEXT NOT NULL
);")

dbExecute(con, "
CREATE TABLE ratings (
  user_id INTEGER,
  song_id INTEGER,
  rating INTEGER CHECK (rating BETWEEN 1 AND 5),
  PRIMARY KEY (user_id, song_id),
  FOREIGN KEY (user_id) REFERENCES users(user_id),
  FOREIGN KEY (song_id) REFERENCES songs(song_id)
);")
```

```{r populate-db}
dbExecute(con, "INSERT INTO users (user_id, name) VALUES
  (1, 'Person 1'), (2, 'Person 2'), (3, 'Person 3'), (4, 'Person 4'), (5, 'Person 5');")

dbExecute(con, "INSERT INTO songs (song_id, title) VALUES
  (1, 'Dine and Dash'), (2, 'Lonely in the Future'), (3, 'Going to Babylon'),
  (4, 'Going Shopping'), (5, \"Liar's Remorse\"), (6, 'Psycho Shit');")


dbExecute(con, "INSERT INTO ratings (user_id, song_id, rating) VALUES
  (1,1,4), (1,2,5), (1,3,5), (1,4,3), (1,5,3), (1,6,2),
  (2,1,3), (2,2,4), (2,3,5), (2,4,3), (2,5,2), (2,6,2),
  (3,1,4), (3,2,5), (3,3,3), (3,4,3),          (3,6,2),
  (4,1,3), (4,2,5), (4,3,3), (4,4,3), (4,5,2), (4,6,2),
  (5,1,3), (5,2,4), (5,3,3),          (5,5,3), (5,6,3);")
```

```{r load-data}

ratings_full <- dbGetQuery(con, "
  SELECT u.name AS listener, s.title AS song, r.rating
  FROM users u
  CROSS JOIN songs s
  LEFT JOIN ratings r
    ON r.user_id = u.user_id AND r.song_id = s.song_id
  ORDER BY u.name, s.title;
")

glimpse(ratings_full)
```

```{r global-baseline}

global_mean <- mean(ratings_full$rating, na.rm = TRUE)
global_mean

user_bias <- ratings_full %>%
  group_by(listener) %>%
  summarize(user_avg = mean(rating, na.rm = TRUE)) %>%
  mutate(user_bias = user_avg - global_mean)
user_bias

song_bias <- ratings_full %>%
  group_by(song) %>%
  summarize(song_avg = mean(rating, na.rm = TRUE)) %>%
  mutate(song_bias = song_avg - global_mean)
song_bias

baseline_estimates <- ratings_full %>%
  left_join(user_bias, by = "listener") %>%
  left_join(song_bias, by = "song") %>%
  mutate(baseline_estimate = global_mean + user_bias + song_bias)

baseline_estimates
```

```{r recommendations}
recommendations <- baseline_estimates %>%
  filter(is.na(rating)) %>%
  select(listener, song, baseline_estimate) %>%
  arrange(listener, desc(baseline_estimate))

recommendations
```

```{r cleanup}
dbDisconnect(con)
```

## Conclusions

The global baseline estimate predicts a rating of 2.61 for Person 3's
unrated song, "Liar's Remorse," and 2.91 for Person 5's unrated song,
"Going Shopping." Both predictions land below the overall average of 3.29,
though for different reasons: Person 3 actually tends to rate songs
slightly above average (+0.11 bias), but "Liar's Remorse" is one of the
two lowest-rated songs overall (-0.79 bias), so the song's weakness pulls
the prediction down despite the listener's generosity. Person 5, on the
other hand, rates close to the group average (-0.09 bias), and "Going
Shopping" is also close to average (-0.29 bias), so that estimate lands
nearer the global mean.

More broadly, "Lonely in the Future" stands out as the strongest song in
the dataset (+1.31 bias), while "Psycho Shit" is the weakest (-1.09 bias)
consistent with what a personalized recommender would also need to
account for, though the global baseline approach intentionally ignores
individual taste patterns in favor of these simple, broad averages.

Future work could compare these baseline estimates against a personalized
method like item item collaborative filtering, to see whether accounting
for individual listener similarity produces meaningfully different (and
more accurate) predictions than this non-personalized approach.